Comprehensive Algorithmic Resilience for Scientific Applications

### 2013-01-22 paper.tex
Fixed Fig 4

This paper presents a comprehensive algorithmic resilience approach that considers all of the above issues.
We demonstrate it on different types of numerical applications: the Alternating Direction Method of Multipliers (ADDR) for Lasso problems~\cite{lasso:2011}, the Hattrick gravitational simulation~\cite{hattrick:2012} and the Digital Room Correction (DRC) acoustic correction application~\cite{drc:2012}.
We show how these three applications can be comprehensively protected from soft faults by adding three different resilience mechanisms to the routines where they spend the most time: algorithmic error checks, replication of critical data structures and checkpoint-restart of individual routines.
We demonstrate via fault injection experiments that although none of these techniques can individually protect applications from soft faults, their combination is highly effective.
Further, we demonstrate how these techniques can be configured to offer the right level of protection at any given fault rate or input type.
More specifically, these configurations can be used to produce a wide range of overheads and result accuracy levels and to tune the probabilistic confidence level that the given overhead or accuracy target will be met.
\greg{Summarize real results} \sui{A tighter algorithmic checker bounds can be enabled at low error rates to completely eliminate erroneous results at a lower overhead, while a more lenient algorithmic checker would allow the user to produce more acceptable results in face of a higher error rate (such as 1e-09).}

@@

Cosmic rays or alpha particles hitting the computer circuits and accumulating a sufficient amount of energy may invert the device logic states from "0" to "1" or from "1" to "0"~\cite{ibm_softerr:1996,baumann:2005}.
These inversions are termed as soft errors or soft faults.
The shrinking processor and memory feature size and the lower threshold voltage coming with relentless manufacturing technology scaling make modern high performance computers are highly vulnerable to soft errors which propagate as the program is being executed.
Prior studies have demonstrated that soft errors occurring in all the important bits that are required for Architecturally Correct Execution (ACE)~\cite{avf:2003} eventually change the program outcome, i.e., manifest errors to the application.
The ACE bits include critical information to keep the program logic such as the program counter register for a committed instruction.
In the past decades, soft errors have brought significant losses on commercial servers and high performance computers such as the recall of Sun servers due to memory errors~\cite{SoftErr:Baumann:2002} or the cache corruptions in the ASCI Q system~\cite{asciQSER:2005}.

@@

The DRC program generates filters for HiFi audio systems to compensate for the reflection of sounds in a room, using impulse response measurements of the audio equipment and considering the positions of the listeners.
The inputs are stored in Pulse Code Modulation (PCM) format, which is an array of 32-bit floating point numbers representing the samples at each sampling time step and the computations are performed using 32-bit precision.
Although the algorithm is divided into many phases the most computationally intensive step consists of the GSL's implementation of the Fast Fourier Transform and an DRC-internal implementation of Finite Impulse Response filter generation.
This program accepts user-controllable configurations to decide the width and number of iterations of FFT transforms it uses to perform various passes on the input file, such as windowing, deconvolution, pre- and post-filtering.
Different configurations would load the CPU for different lengths of time, resulting in the program generating inputs of different sizes \sui{to the FFT routines.}.
\greg{What inputs?}

\subsection{Hattrick}
\label{sec:apps:hattrick}
The Hattrick application simulates the motion of bodies under the effects of gravity using 64-bit precision.
It is designed to help discover extra-solar planets by inferring their existence from Transit Timing Variations, where the difference between the times when a planet passes in front of its host star is used to infer the existence and properties of other planets in the system.
This program spends most of its execution time in the GSL solver for Ordinary Differential Equations to solver the system's equations of motion.
\greg{What inputs?} \sui{Input files are four different planetary systems. The first, third and fourth input files simulate one star and one planet and the second simulates one star and two planets. The third and fourth inputs are the same planetary system with different lengths of simulation time.}

@@

Each algorithm used by the above application was enhanced with an algorithm-specific checker that validated its results by exploiting some algorithmic identity.
Each checker computes some aggregate property of the algorithm's results using two different algorithm to produce vectors $v$ and $v'$ and their relative difference by more than some threshold $\tau$, the checker declares that an error has occurred.
The relative difference is computed as $\frac{\left|| v-v' \right||}{max(\left||v\right||, \left||v'\right||)}$ \greg{is this the right formula???} \sui{yes}.
Each error detection causes the routine's execution to be restarted using the rollback mechanism described in Section~\ref{sec:res_tech:cr}.

\paragraph{Matrix-matrix multiplication (MMM)}
The MMM operation $A \cdot B$ is checked using a matrix vector multiplication (MVM), making use of this identity: $(A \cdot B) \cdot x = A \cdot (B \cdot x)$, where $x$ is an error-checking vector.
Both sides compute the sums of the columns of matrix $A \cdot B$ using a single matrix-vector multiplication on the left and two such operations on the right.
Most errors in the computation of $A \cdot B$ will affect the sums of its columns and will thus be detected.
This check is efficient because the checker is asymptotically faster than the MMM algorithm, with MVM taking $O(n^2)$ time and MMM $O(n^3)$ time.
Since we make no assumptions about the distribution of data in the matrices being multiplied, we set the check vector $x$ to contain all 1's.

@@

When errors are detected we employ a recovery method based on checkpoint-restart, where routine state is recorded at key locations and whenever an algorithmic or memory error is detected we roll back to that point in its execution.
Rollbacks are implemented via the \texttt{sigsetjmp}/\texttt{siglongjmp} routines.
\texttt{sigsetjmp} record the application's execution state (registers and stack pointer) at the time of the checkpoint.
On error detection \texttt{siglongjmp} returns the application to the same execution point, unrolling any function calls between the checkpoint and the error detection.
This method is useful for rolling back to checkpoints recorded in the same function or in a caller function.
When execution returns to the \texttt{sigsetjmp} call the checkpointed portions of the routine's state are recovered and execution repeats.
% The overhead of this API call is having to copy the processor state into memory, therefore installing them between computation-intensive method calls adds to negligible performance degradation.

Checkpoint data is protected using two different mechanisms.
Pointers and scalars are replicated such that the checkpoint contains two copies of each.
Arrays of floating point numbers are protected via a block-checksum-based error correcting code..
By treating the input array as a matrix and comparing row sums and column sums of the submatrices of that matrix, we are able to detect the existence of errors and fix a fraction of the errors.
\sui{This algorithm takes in an array of numbers, divide it into blocks of $N \cdot N$ numbers, computes and stores the row sums and column sums that are used for recovery. It is capable of correcting one corrupt element that does not share the same row or column with another element.}
The protection mechanism has the following parameters:
\begin{itemize}
\item{\texttt{N}, the size of the error correcting sum block.}
\item{\texttt{ECC\_ECC}, whether or not the error correcting code itself should be corrected.}
\end{itemize}
This mechanism is able to correct up to $1/{n^2}$ of the elements in the input array.
\greg{Which code is being used here? You just say that it is based on block-checksums}.

The structure of each routine determines how checkpointing is deployed.
All routines except RK are monolithic code blocks that take a relatively large amount of data as input, operate on it and then return.
In contrast, RK takes in very little input and operates as a long sequence of steps, each of which involves a call to a user-provided derivative routine.
As such, we checkpoint the state of non-RK routines immediately after the call to the routine and include in this checkpoint the routine's arguments and contents of the floating point input buffers \greg{is this correct?} \sui{Yes}.
The one exception is the FIR routine, which takes no input and thus requires only the processor's state to be checkpointed..
Because RK's checkpoints are very small and it is more sensitive to errors, it is checkpointed periodically during its execution with a fixed number of steps between checkpoints.
As our experiments \greg{where?} \sui{This should be ready tomorrow} show, the choice of the RK checkpoint period significantly affects its efficiency with longer periods suffering from more overhead due to rollbacks and short period spending more time checkpointing.

@@

\begin{figure}[ht!]
%\vspace{-20pt}
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_linalg}
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_fft}
%\vspace{-5pt}
\caption{Overhead of resilience techniques when errors are not injected \greg{Change y-axis name to say overhead rather than ratio} \sui{Done}}
\label{fig:routine_detect_ovhd}
\end{figure}

%Experiment 1: Measure performance degradation of using error detectors as input sizes increase. In case of RK if we can run a version that doesn't do the multiple step sizes, do so but if this is difficult then don't worry about it.
We evaluate the costs of the above algorithmic error checkers by measuring their costs and effectiveness when running at different error rates on different types of inputs.
Figure \ref{fig:routine_detect_ovhd} measures the overheads of employing the algorithmic error detectors when no errors are injected, reporting the average of \greg{???} \sui{100} runs.
The linear algebra routines were executed on $nxn$ square matrices where $n$ varies from 100 to 1000.
FFT ran on inputs vectors with 128K to 4M entries and RK ran on an integral with intervals $[0, t_1]$ where $t_1 \in \{10, 15, 20, 25, 30, 50, 75, 100\}$. \greg{What function was being computed by RK?} \sui{The ODE system consists of a second-order nonlinear Van der Pol oscillator equation $x''(t)+\mu x'(t)(x(t)^2-1)+x(t)=0$}
As our experiments show, the cost of algorithmic error detection is low and drops with increasing input sizes.

\begin{figure}[ht!]
%\vspace{-20pt}
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Overall_Breakdown.png}
%\vspace{-5pt}
\caption{Overhead of resilience techniques when no errors are injected \sui{(Bars) and when errors are injected (points). (The SYRK overhead is from the old algorithm (the O($n^2 logn$) version; will update today! }}
\label{fig:routine_all_ovhd}
\end{figure}

Figure~\ref{fig:routine_all_ovhd} shows the overhead of all resilience techniques with no fault injection.
The data shows that checkpointing consumes as much, if not more, time than error detection but that this cost grows more slowly than the routines' computational cost.
In routines MV and RK the inputs are sufficiently small that checkpointing has negligible cost.
Howveer, because MV runs quickly and because RK takes many checkpoints (\greg{what's the period in this experiment?} \sui{if we take a checkpoint each time RK is called, we can end up taking millions of checkpoints, several magnitudes more than we do for MV and for other linear algebra routines and FFT)}), the cost of saving processor state is small but not negligible.

Finally, Figure~\ref{fig:routine_exp_exec} documents the overhead of resilience when errors are injected into application execution, accounting for the slowdowns due to the resilience mechanisms as well as rollbacks due to error detections.
Each routine's graph shows $\frac{\texttt{Time with Errors}}{\texttt{Time No Errors}}$, where the \texttt{Time with Errors} is average increase execution time of a resilient or non-resilient version of the routine when running on the same range of inputs as above, and \texttt{Time No Errors} is the execution time of the non-resilient version on the same inputs when no errors are injected.
The graph for each routine reports this metric for all of the above input sizes, error injection rates between 1e-8 and \sui{(1e-8 is the highest error rate that we can have for most runs to complete) and 1e-10 (is the lowest error rate we would observe errors from mosts of the runs) } \greg{confirm!} and all resilience configurations (no resilience and resilience with detection thresholds between 1e-6 and 1e-8), as the average time of \greg{???} runs \sui{100 runs per configuration}.
The data points within each figure are organized horizontally according to the average number of errors injected into runs with the given input size and at the given injection rate. \greg{Do the error counts include errors after rollbacks?} \sui{No, because the error count is the number of errors for Non-FT runs with errors. Since there is no FT and no checkpointing in these configurations, the number doesn't include errors after rollbacks. }
This average error count is shown on the x-axis.
All runs with the same resilience configuration are joined together into a line to highlight its properties.

The data shows \greg{???} \sui{Soft errors may cause the routines to run longer due to rollback in attempt to correct the errors. The higher the error rate, the more likely rollback is to occur. For example, the Cholesky Decomposition routine with input size 100, when exposed to an error rate of 1e-06, would run on average twice as long as when no fault is injected. That means in this case, every one of the runs tend to have 100\% overhead. In some cases the algorithm would give up retrying after the fifth time because it is unable to recover from the corrupt inputs. For example, for FFT exposed at a fault rate of 1e-08, the average overhead, as computed by dividing the time spent in the fault-tolerance codes by the baseline running time, decreases, because more runs tend to early-terminate when too many elements in the input have been corrupted.}

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp2_1_Example.png}
\caption{Error magnitude histogram showing the effectiveness of the error detector \greg{Why is FFT not improving with tighter threshold? Why is MM differently vulnerable than SYRK?} \sui{Reason for FFT: 1. Errors smaller than a certain value cannot be captured by the checker, even at low error rates, so this should be the fundamental limitation of the checker. 2. We actually are seeing reduction in large errors, but those large errors account for such a small fraction that they are not obviously visible on the graph. For the difference between MM and SYRK: MM is more vulnerable therefore the difference between ``corrected'' and ``not corrected'' would look more noticable.}}
\label{fig:algo_err_dist}
\end{figure}

>>>>

This paper presents a comprehensive algorithmic resilience approach that considers all of the above issues.
We demonstrate it on different types of numerical applications: the Alternating Direction Method of Multipliers (ADDR) for Lasso problems~\cite{lasso:2011}, the Hattrick gravitational simulation~\cite{hattrick:2012} and the Digital Room Correction (DRC) acoustic correction application~\cite{drc:2012}.
We show how these three applications can be comprehensively protected from soft faults by adding three different resilience mechanisms to the routines where they spend the most time: algorithmic error checks, replication of critical data structures and checkpoint-restart of individual routines.
We demonstrate via fault injection experiments that although none of these techniques can individually protect applications from soft faults, their combination is highly effective.
Further, we demonstrate how these techniques can be configured to offer the right level of protection at any given fault rate or input type.
More specifically, these configurations can be used to produce a wide range of overheads and result accuracy levels and to tune the probabilistic confidence level that the given overhead or accuracy target will be met.
\greg{Summarize real results} \sui{A tighter algorithmic checker bounds can be enabled at low error rates to completely eliminate erroneous results at a lower overhead, while a more lenient algorithmic checker would allow the user to produce more acceptable results in face of a higher error rate (such as 1e-09). Through high and low error rates, we correct more than 50\% of segmentation faults and reduce the error in the outputs by magnitudes by paying lower than 20\% overheads for linear algebra, FFT and FIR filter generation and a reasonable overhead for the integrator.}

@@

Cosmic rays or alpha particles hitting the computer circuits and accumulating a sufficient amount of energy may invert the device logic states from "0" to "1" or from "1" to "0"~\cite{Ziegler:1996:TCR:226354.226356,baumann:2005}.
These inversions are termed as soft errors or soft faults.
The shrinking processor and memory feature size and the lower threshold voltage coming with relentless manufacturing technology scaling make modern high performance computers are highly vulnerable to soft errors which propagate as the program is being executed.
Prior studies have demonstrated that soft errors occurring in all the important bits that are required for Architecturally Correct Execution (ACE)~\cite{avf:2003} eventually change the program outcome, i.e., manifest errors to the application.
The ACE bits include critical information to keep the program logic such as the program counter register for a committed instruction.
In the past decades, soft errors have brought significant losses on commercial servers and high performance computers such as the recall of Sun servers due to memory errors~\cite{baumann:2005} or the cache corruptions in the ASCI Q system~\cite{asciQSER:2005}.

@@

The DRC program generates filters for HiFi audio systems to compensate for the reflection of sounds in a room, using impulse response measurements of the audio equipment and considering the positions of the listeners.
The inputs are stored in Pulse Code Modulation (PCM) format, which is an array of 32-bit floating point numbers representing the samples at each sampling time step and the computations are performed using 32-bit precision.
Although the algorithm is divided into many phases the most computationally intensive step consists of the GSL's implementation of the Fast Fourier Transform and an DRC-internal implementation of Finite Impulse Response filter generation.
This program accepts user-controllable configurations to decide the width and number of iterations of FFT transforms it uses to perform various passes on the input file, such as windowing, deconvolution, pre- and post-filtering.
The input used in this paper is an audio file of size 768 kilobytes.

\subsection{Hattrick}
\label{sec:apps:hattrick}
The Hattrick application simulates the motion of bodies under the effects of gravity using 64-bit precision.
It is designed to help discover extra-solar planets by inferring their existence from Transit Timing Variations, where the difference between the times when a planet passes in front of its host star is used to infer the existence and properties of other planets in the system.
This program spends most of its execution time in the GSL solver for Ordinary Differential Equations to solver the system's equations of motion.
Input files are four different planetary systems. The first, third and fourth input files simulate one star and one planet and the second simulates one star and two planets. The third and fourth inputs are the same planetary system with different lengths of simulation time.

@@

Each algorithm used by the above application was enhanced with an algorithm-specific checker that validated its results by exploiting some algorithmic identity.
Each checker computes some aggregate property of the algorithm's results using two different algorithm to produce vectors $v$ and $v'$ and their relative difference by more than some threshold $\tau$, the checker declares that an error has occurred.
The relative difference is computed as $\frac{\left|| v-v' \right||}{max(\left||v\right||, \left||v'\right||)}$.
Each error detection causes the routine's execution to be restarted using the rollback mechanism described in Section~\ref{sec:res_tech:cr}.

\paragraph{Matrix-matrix multiplication (MMM)}
The MMM operation $A \cdot B$ is checked using a matrix vector multiplication (MVM), making use of this identity: $(A \cdot B) \cdot x = A \cdot (B \cdot x)$, where $x$ is an error-checking vector.
Both sides compute the sums of the columns of matrix $A \cdot B$ using a single matrix-vector multiplication on the left and two such operations on the right.
Most errors in the computation of $A \cdot B$ will affect the sums of its columns and will thus be detected.
This check is efficient because the checker is asymptotically faster than the MMM algorithm, with MVM taking $O(n^2)$ time and MMM $O(n^3)$ time.
Since this is dense martix matrix multiplication, we set the check vector $x$ to contain all 1's.

@@

When errors are detected we employ a recovery method based on checkpoint-restart, where routine state is recorded at key locations and whenever an algorithmic or memory error is detected we roll back to that point in its execution.
Rollbacks are implemented via the \texttt{sigsetjmp}/\texttt{siglongjmp} \cite{sigsetjmp:1997} routines.
\texttt{sigsetjmp} record the application's execution state (registers and stack pointer) at the time of the checkpoint.
On error detection \texttt{siglongjmp} returns the application to the same execution point, unrolling any function calls between the checkpoint and the error detection.
This method is useful for rolling back to checkpoints recorded in the same function or in a caller function.
When execution returns to the \texttt{sigsetjmp} call the checkpointed portions of the routine's state are recovered and execution repeats.
% The overhead of this API call is having to copy the processor state into memory, therefore installing them between computation-intensive method calls adds to negligible performance degradation.

Checkpoint data is protected using two different mechanisms.
Pointers and scalars are replicated such that the checkpoint contains two copies of each.
Arrays of floating point numbers are protected via a block-checksum-based error correcting code..
By treating the input array as a matrix and comparing row sums and column sums of the submatrices of that matrix, we are able to detect the existence of errors and fix a fraction of the errors.
This algorithm takes in an array of numbers, divide it into blocks of $N \cdot N$ numbers and for each block it computes and stores the row sums and column sums that are used for recovery.
The protection mechanism has one parameters,\texttt{N}, the size of the error correcting sum block.

The structure of each routine determines how checkpointing is deployed.
All routines except RK are monolithic code blocks that take a relatively large amount of data as input, operate on it and then return.
In contrast, RK takes in very little input and operates as a long sequence of steps, each of which involves a call to a user-provided derivative routine.
As such, we checkpoint the state of non-RK routines immediately after the call to the routine and include in this checkpoint the routine's arguments and contents of the floating point input buffers.
The one exception is the FIR routine, which takes no input and thus requires only the processor's state to be checkpointed..
Because RK's checkpoints are very small and it is more sensitive to errors, it is checkpointed periodically during its execution with a fixed number of steps between checkpoints.
As our experiments in \ref{sec:res_tech:eval:rk} show, the choice of the RK checkpoint period significantly affects its efficiency with longer periods suffering from more overhead due to rollbacks and short period spending more time checkpointing.

@@

\begin{figure}[ht!]
%\vspace{-20pt}
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_linalg}
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_fft}
%\vspace{-5pt}
\caption{Overhead of resilience techniques when errors are not injected \greg{Change y-axis name to say overhead rather than ratio}}
\label{fig:routine_detect_ovhd}
\end{figure}

%Experiment 1: Measure performance degradation of using error detectors as input sizes increase. In case of RK if we can run a version that doesn't do the multiple step sizes, do so but if this is difficult then don't worry about it.
We evaluate the costs of the above algorithmic error checkers by measuring their costs and effectiveness when running at different error rates on different types of inputs.
Figure \ref{fig:routine_detect_ovhd} measures the overheads of employing the algorithmic error detectors when no errors are injected, reporting the average of 100 runs.
The linear algebra routines were executed on $nxn$ square matrices where $n$ varies from 100 to 1000.
FFT ran on inputs vectors with 128K to 4M entries and RK ran on an integral with intervals $[0, t_1]$ where $t_1 \in \{10, 15, 20, 25, 30, 50, 75, 100\}$. The ODE system consists of a second-order nonlinear Van der Pol oscillator equation $x''(t)+\mu x'(t)(x(t)^2-1)+x(t)=0$.
As our experiments show, the cost of algorithmic error detection is low and drops with increasing input sizes.

\begin{figure}[ht!]
%\vspace{-20pt}
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Overall_Breakdown.png}
%\vspace{-5pt}
\caption{Overhead of resilience techniques when no errors are injected and when errors are injected (points).}
\label{fig:routine_all_ovhd}
\end{figure}

Figure~\ref{fig:routine_all_ovhd} shows the overhead of all resilience techniques with no fault injection.
The data shows that checkpointing consumes as much, if not more, time than error detection but that this cost grows more slowly than the routines' computational cost.
In routines MV and RK the inputs are sufficiently small that checkpointing has negligible cost.
Howveer, because MV runs quickly and because RK takes many checkpoints ($9000 \cdot input size)$), the cost of saving processor state is small but not negligible.

%Finally, Figure~\ref{fig:routine_exp_exec} documents the overhead of resilience when errors are injected into application execution, accounting for the slowdowns due to the resilience mechanisms as well as rollbacks due to error detections.

\sui{(Describe Fig.2 when faults are injected)}

%Each routine's graph shows $\frac{\texttt{Time with Errors}}{\texttt{Time No Errors}}$, where the \texttt{Time with Errors} is average increase execution time of a resilient or non-resilient version of the routine when running on the same range of inputs as above, and \texttt{Time No Errors} is the execution time of the non-resilient version on the same inputs when no errors are injected.
%The graph for each routine reports this metric for all of the above input sizes, error injection rates between 1e-8 and \sui{(1e-8 is the highest error rate that we can have for most runs to complete) and 1e-10 (is the lowest error rate we would observe errors from mosts of the runs) } \greg{confirm!} and all resilience configurations (no resilience and resilience with detection thresholds between 1e-6 and 1e-8), as the average time of \greg{???} runs \sui{100 runs per configuration}.
The data points within each figure are organized horizontally according to the average number of errors injected into runs with the given input size and at the given injection rate. \greg{Do the error counts include errors after rollbacks?} \sui{No, because the error count is the number of errors for Non-FT runs with errors. Since there is no FT and no checkpointing in these configurations, the number doesn't include errors after rollbacks. }
This average error count is shown on the x-axis.
All runs with the same resilience configuration are joined together into a line to highlight its properties.

\sui{These are about Figure 2}The data shows \greg{???} \sui{Soft errors may cause the routines to run longer due to rollback in attempt to correct the errors. The higher the error rate, the more likely rollback is to occur. For example, the Cholesky Decomposition routine with input size 100, when exposed to an error rate of 1e-06, would run on average twice as long as when no fault is injected. That means in this case, every one of the runs tend to have 100\% overhead. In some cases the algorithm would give up retrying after the fifth time because it is unable to recover from the corrupt inputs. For example, for FFT exposed at a fault rate of 1e-08, the average overhead, as computed by dividing the time spent in the fault-tolerance codes by the baseline running time, decreases, because more runs tend to early-terminate when too many elements in the input have been corrupted.}

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp2_1_Example.png}
\caption{Error magnitude histogram showing the effectiveness of the error detector}
\label{fig:algo_err_dist}
\end{figure}
